Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-11-15 manuscript/manuscript.tex
more literature changes

In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see~\cite{dror2012biomolecular} for a review).

@@

For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview,
while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

\begin{itemize}
	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
	\item Clustering -- This is the step at which the state decomposition occurs.
	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
\end{itemize}

@@

Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space,
which typically contains many degrees of freedom,
to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff
(note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
Discrete jumps between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

Additional technical challenges for large systems include high demands on memory and computation time;
we explain how to deal with those in the tutorials (notebook~01).

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in notebook~08.

@@

Conventional Markov state modeling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram},
we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange
with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}. 
MEMMs are implemented in PyEMMA.

>>>>

In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).

@@

For those seeking further resources, the recent perspective ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a timeline of methods advances with relevant citations,
while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and provides the underlying mathematics in detail.
Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,mdtutorial},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

\begin{itemize}
	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
	\item Clustering -- This is the step at which the state decomposition occurs.
	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
\end{itemize}

@@

Subsequently, we perform TICA~\cite{tica,tica3,kinetic-maps} in order to reduce the dimension from the feature space,
which typically contains many degrees of freedom,
to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff
(note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
Discrete jumps between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

@@

Additional technical challenges for large systems include high demands on memory and computation time;
we explain how to deal with those in the tutorials (notebook~01).
More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in notebook~08.

@@

Conventional Markov state modeling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram},
we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange
with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}. 
MEMMs are implemented in PyEMMA.
Since the many publications associated with the development of these methods are beyond the scope of this tutorial, we refer the reader to Sec.~8.3 of Ref.~\cite{msm-brooke} and the references therein.
